---
title: "Notes"
author: "Ryan Heslin"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
  pdf_document:
    highlight_style: "kate"
    df_print: "kable"
---

 <!-- Standard custom LaTeX commands -->
\newcommand{\abcd}{\begin{bmatrix}a&b\\
c&d\end{bmatrix}}
\newcommand{\m}[1]{\begin{bmatrix}#1\end{bmatrix}}

\newcommand{\vect}[1]{\begin{pmatrix}#1\end{pmatrix}}

\newcommand{\meq}[1]{\begin{split}#1\end{split}}

\newcommand{\bym}[1]{#1\times{m}}

\newcommand{\nby}[1]{n\times{#1}}

\newcommand{\subsp}[2]{\Bigg\{\begin{bmatrix}#1\end{bmatrix}:#2\Bigg\}}

\newcommand{\proj}[2]{\text{proj}_#1(#2)}

\newcommand{\refl}[2]{\text{refl}_#1(#2)}

\newcommand{\sumn}{\sum_{i=1}^n}

<!-- % 1: term 1 -->
<!-- % 2: subscript 1 -->
<!-- % 3: term 2 -->
<!-- % 4: subscript 2 -->
<!-- % 5. operation -->
\newcommand{\dotsn}[5]{#1_{1}#3_{1}#5{#1}_{2}#3_{2}{#5}\dots{#5}#1_{#2}#3_{#4}}

```{r setup, include=FALSE}

knitr::opts_chunk$set(
  echo = TRUE,
  comment = "",
  fig.pos = "",
  warning = FALSE,
  tidy = TRUE,
  fig.align = "center",
  highlight = TRUE
)

library(tidyverse)
library(rlang)
```



```{r}
raw <- read_csv("../data/World_MO_Hurricane_Survival.csv")

```


```{r}
sort_cat <- function(fac){
  
  new <- str_extract(unique(fac), "^\\d+") %>% 
    as.numeric() %>%
    order()
  unique(fac)[new]
}

cleaned <- raw %>% rename_with(str_to_lower) %>% 
  mutate(across(where(is.character), str_to_title),
         across(c(dist_coast, mntly_travel), ~str_remove(.x, "\\s Km")),
         across(c(dist_frm_coast, mntly_travel), ~factor(.x, ordered = TRUE, levels = sort_cat(.x)))) %>% 
           mutate(salary = factor(salary, levels = c("0-100k", "100k-300k", "500k-700k", "700k-1 Million", "1m-2 Million"), ordered = TRUE ),
                edu_data = factor(edu_data, levels = c("Uneducated", "High-School", "Graduate", "Post-Graduate")))
cleaned$class <- as.factor(c(X = "Survived", Y = "Died")[cleaned$class]) 



```



```{r}
invisible(map_if(cleaned, is.factor, plot))
```

```{r}
cleaned %>% select(where(is.factor), -class) %>%
  colnames() %>%
  map(
    ~ cleaned %>% group_by(!!sym(.x)) %>%
      summarize(survived = mean(class == "Survived")) %>% ggplot(aes(
        x = .data[[.x]], y = survived, fill = .data[[.x]]
      )) +
      geom_col()
  )
```

